---
title: "RepData_Project1"
output: html_document
date: "2026-01-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## Loading and preprocessing the data

Read in the csv:

```{r}
Data <- read.csv("C:\\Users\\kroberts\\Downloads\\repdata_data_activity\\activity.csv")
```

## What is the mean total number of steps taken per day?

Add columns for mean and median steps, create histograms, report values:

```{r}
daily_sum_steps <- data.frame()
uniquedates <- unique(Data$date)

for (i in uniquedates){
  dailydata <- Data[which(Data$date == i),]
  daily_sum_steps <- rbind(daily_sum_steps,data.frame(i,sum(dailydata$steps, na.rm = TRUE)))
}

colnames(daily_sum_steps) <- c("Date","Sum")

ggplot(daily_sum_steps, aes(x = `Sum`)) +
 geom_histogram(color = "black", fill = "gold", ) +
  labs(
    title = "Histogram of Total Number of Steps",
    x = "Daily Total Steps",
    y = "Count of Days"
  ) +
  theme_classic(paper = "cornsilk1")

MeanSteps <- mean(daily_sum_steps$Sum, na.rm = TRUE)
print(paste0("Mean Total Steps per Day = ", MeanSteps))
MedianSteps <- median(daily_sum_steps$Sum, na.rm = TRUE)
print(paste0("Median Total Steps per Day = ",MedianSteps))

```

## What is the average daily activity pattern?

```{r}
# Compute average steps per 5-minute interval (ignore NA values)
avg_steps <- Data %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm = TRUE))

ggplot(avg_steps, aes(x = interval, y = mean_steps)) +
  geom_line(colour = "red")+
  geom_point()+
  labs(title = "Average Daily Activity Pattern",
       x = "5-minute Interval",
       y = "Average Number of Steps") +
  theme_classic(paper = "cornsilk1")

# Identify the interval with the maximum average steps
max_interval <- avg_steps$interval[which.max(avg_steps$mean_steps)]
max_steps <- max(avg_steps$mean_steps)

# Print results
print(paste0("5-minute interval with maximum average steps:", max_interval))
  
```

## Inputing missing values

```{r}
#Calculate number of NA's
numna <- length(which(is.na(Data$steps)))
print(paste0("Total Number of NA values = ",numna))

#Fill in missing values in dataset with mean for 5-minute interval
filled_Data <- Data
for (i in 1:nrow(filled_Data)){
  if (is.na(filled_Data[i,"steps"])){
    intv <- filled_Data[i,"interval"]
    filled_Data[i,"steps"] <- avg_steps[which(avg_steps$interval == intv), "mean_steps"]
  }
}

#Find total steps per day for filled in data
daily_sum_steps2 <- data.frame()
for (i in uniquedates){
  dailydata2 <- filled_Data[which(filled_Data$date == i),]
  daily_sum_steps2 <- rbind(daily_sum_steps2,data.frame(i,sum(dailydata2$steps, na.rm = TRUE)))
}
colnames(daily_sum_steps2) <- c("Date","Sum")

#histogram of total steps per day for filled in dataset
ggplot(daily_sum_steps2, aes(x = Sum))+
  geom_histogram(color = "black", fill = "gold", ) +
  labs(
    title = "Histogram of Total Number of Steps (NA's replaced)",
    x = "Daily Total Steps",
    y = "Count of Days"
  ) +
  theme_classic(paper = "cornsilk1")

MeanSteps2 <- mean(daily_sum_steps2$Sum, na.rm = TRUE)
print(paste0("Mean Total Steps per Day = ", MeanSteps2))
MedianSteps2 <- median(daily_sum_steps2$Sum, na.rm = TRUE)
print(paste0("Median Total Steps per Day = ",MedianSteps2))
```
The mean/median of the dataset with filled NA's are higher than the mean/median of the initial dataset. 
The estimates of total daily steps increased due to inputing missing data.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
#determine weekend/weekday
weekdays_data <- weekdays(as.POSIXct(filled_Data$date, format = "%Y-%m-%d"), abbreviate = FALSE)
weekday_col <- data.frame()
for (x in weekdays_data){
  if(grepl("Sat|Sun",x)){
    day <- "weekend"
  } else {
    day <- "weekday"
  }
  weekday_col <- rbind(weekday_col,day)
}
colnames(weekday_col) <- "Weekdays"
filled_Data <- cbind(filled_Data,weekday_col)

# #find avg steps per day
# avg_steps2 <- filled_Data %>%
#   group_by(interval) %>%
#   summarise(mean_steps = mean(steps))
# 
# #Create panel plot
# ggplot(filled_Data, aes(x = "interval", y = "steps")) +
#   geom_point() +
#   facet_wrap(~Weekdays)

# Find average steps per interval per day type
avg_steps2 <- filled_Data %>%
  group_by(interval, Weekdays) %>%
  summarise(mean_steps = mean(steps), .groups = "drop")

# Create panel plot (type = "l" equivalent is geom_line())
ggplot(avg_steps2, aes(x = interval, y = mean_steps)) +
  geom_line(colour = "red") +
  geom_point() +
  facet_wrap(~Weekdays, ncol = 1) +
  labs(
    title = "Average Daily Activity Pattern",
    x = "5-minute Interval",
    y = "Average Number of Steps"
  ) +
  theme_classic(paper = "cornsilk1")

```